---
title: "Port Phillip Bay seawater temps"
output:
  html_document: 
       theme: cosmo
  pdf_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r include=FALSE}
library(tidyverse)
library(ggsci)
library(rerddap)
library(giscoR)
library(patchwork)
```


The temperature records are satellite sea surface temperatures sourced from [NOAA](https://upwell.pfeg.noaa.gov/erddap/griddap/nesdisGeoPolarSSTN5NRT.html) and presented as °C in the dataset *NOAA_DHW*. The data come from 0.05° × 0.05° grid cells, corresponding to approximately 5 km × 5 km. The data can be viewed interactively, but examining multi-year time series or making site comparisons is not straightforward.

The grid cell closest to Williamstown beach is centred on 144.875°E, 37.875°S. Two other snorkelling and diving sites are included: the south-east corner of the bay (Blairgowrie, Rye, and surrounds) at 144.775°E, 38.375°S, and a location just outside Western Port near Flinders Pier at 145.025°E, 38.475°S.

```{r params, include=FALSE}
#Downloading using erddap protocol
#Set basic parameters for data download
sstInfo <- info('NOAA_DHW', url = "https://upwell.pfeg.noaa.gov/erddap/")
#Retrieves basic structure of NOAA dataset
```

```{r include=FALSE}
latx = c(-37.8, -38.525)    #Set bounding box for download
lonx = c(144.35, 145.3)
```

```{r include=FALSE, eval=FALSE}
#Use this chunk to make new clean copy of stored dataframe "bay"
#Normally just update saved file each time
timex = c("2003-01-01", "last")     #Start download from 2013
bay <- griddap(sstInfo, latitude = latx, longitude = lonx, time = timex, fmt= "csv", fields = 'CRW_SST')
save(bay, file="bay.Rdata")  #Saved version from last run
```

```{r getstart, include=FALSE}
load("bay.Rdata")     #Load saved file
start <- as.Date(max(bay$time))+1   #Get next day after data in saved file
```

```{r getdata, include=FALSE}
timex = c(as.character(start), "last")   #Used to download only new data, rather than large data set each time
baynew <- griddap(sstInfo, latitude = latx, longitude = lonx, time = timex, fmt= "csv", fields = 'CRW_SST')
bay <- bind_rows(bay, baynew)  # Append to saved file
bay <- unique(bay)
save(bay, file="bay.Rdata")  #Save new version
```


```{r tidy, include=FALSE}
bay <- bay |>
  mutate(
    sst      = crw_sst,
    date     = with_tz(ymd_hms(time), tzone = "Australia/Melbourne"),  # Convert to UTC+10
    year     = year(date),                          # Year variable for grouping
    year_chr = as.character(year),                  # Character year for multi-year plots
    day      = yday(date),                          # Day of year
    day_plot = as.Date("2019-12-31") + yday(date),  # Variable for m/d plots
    summer   = year + (month(date) >= 7),           # Grouping by summer
    myLabel  = case_when(                           # Label for 1st of each month
      day(day_plot) == 1L ~ format(day_plot, "%b-%d"),
      TRUE               ~ NA_character_
    )
  )
latest_time <- max(bay$date)
latest      <- date(latest_time)
```


```{r boundaries, include=FALSE}
#For convenience, set 3 desired locations as variables:
wlon <- 144.875
wlat <- -37.875
blon <- 144.775
blat <- -38.375
flon <- 145.025
flat <- -38.475

#Define year to be plotted and temperature ranges for y axis
# For Port Phillip Bay, force plots to useful temperature range (uses all of y-axis). 
# For viewing data from a single specific year, set value of pyear, rather than using formula below.
nyear <- max(bay$year)    #Current year in downloaded data
pyear <- nyear-1    #Set penultimate year
syear <- 2014  #Starting year for plots
tmax <- ceiling(max(bay$sst, na.rm = TRUE))
tmin <- floor(min(bay$sst, na.rm = TRUE))
wlatest <- bay |> filter(time == max(time), longitude == wlon, latitude == wlat)
wtemp   <- paste(format(latest, "%A, %B %d, %Y"), round(wlatest$sst, digits = 1), sep = ": ")

```

## Williamstown time series

#### Current year, with last year for an immediate comparison

```{r jawbonerecent, include=FALSE}
p1 <- ggplot(
    data = bay |> filter(year >= pyear, longitude == wlon, latitude == wlat),
    aes(x = day_plot, y = sst, group = year_chr, colour = year_chr)
  ) +
  geom_line() +
  geom_point(
    data = bay |> filter(year > pyear, longitude == wlon, latitude == wlat, time == max(time)),
    aes(x = day_plot, y = sst)
  ) +
  scale_y_continuous(breaks = seq(tmin, tmax, 1), limits = c(tmin, tmax)) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(y = "C", x = NULL, title = "Sea surface temperature, Williamstown", caption = wtemp) +
  theme_minimal() +
  scale_color_aaas(labels = c("Last year", "This year")) +
  theme(legend.title = element_blank())
```

```{r jawbonesummer, include=FALSE}
df <- bay |>
  filter(summer >= pyear, longitude == wlon, latitude == wlat) |>
  group_by(summer) |>
  mutate(order = row_number()) |>
  ungroup()
x_break <- df$order[!is.na(df$myLabel)][1:12]
x_label <- df$myLabel[x_break]
p1a <- ggplot(data = df, aes(x = order, y = sst, group = as.factor(summer), colour = as.factor(summer))) +
  geom_line() +
  geom_point(
    data = df |> filter(summer > pyear, time == max(time)),
    aes(x = order, y = sst)
  ) +
  scale_y_continuous(breaks = seq(tmin, tmax, 1), limits = c(tmin, tmax)) +
  scale_x_continuous(breaks = x_break, labels = x_label) +
  labs(y = "C", x = NULL, title = "Sea surface temperature, Williamstown", caption = wtemp) +
  theme_minimal() +
  scale_color_aaas(labels = c("Last summer", "This summer")) +
  theme(axis.text.x = element_text(angle = 90), legend.title = element_blank())
```

```{r}
# Choose plot to do based on time of year; use Jan-Dec from April to Aug, Jul-June at other times of year
if(month(latest_time)>3 & month(latest_time)<9) {
  p1
} else {
  p1a
}
```

#### Multi-year plots

The first figure shows the current year with all years from 2014 onwards in the background; the second extends this to include all years since 2002. The third figure plots the current year against 10- and 20-year averages.

```{r multipanel, include=FALSE}
#Data goes from 2002 to present. Having 23 years' data is a bit spaghetti-like. Two options are to plot fewer years and/or emphasise the current year more. Do the first option by specifying a value for syear.
#Do multi-year plot, then overlay data for chosen year (pyear) with heavier line.
jawbone_all <- bay |> filter(longitude == wlon, latitude == wlat)
p2 <- ggplot(data = jawbone_all |> filter(year >= syear), aes(x = day_plot, y = sst)) +
  geom_line(linewidth = 0.25) +
  scale_y_continuous(breaks = seq(tmin, tmax, 2)) +
  scale_x_date(date_breaks = "2 month", date_labels = "%b") +
  labs(y = "C", x = NULL, title = "Daily surface temperature each year") +
  theme_classic() +
  theme(
    legend.title = element_blank(),
    legend.text  = element_text(size = 5),
    legend.spacing.y = unit(.25, "cm")
  ) +
  guides(fill = guide_legend(byrow = TRUE)) +
  facet_wrap(vars(year))

p2c <- ggplot(jawbone_all, aes(x = day_plot, y = sst)) +
  geom_line(linewidth = 0.25) +
  scale_y_continuous(breaks = seq(tmin, tmax, 2)) +
  scale_x_date(date_breaks = "2 month", date_labels = "%b") +
  labs(y = "C", x = NULL, title = "Daily surface temperature each year") +
  theme_classic() +
  theme(
    legend.title = element_blank(),
    legend.text  = element_text(size = 5),
    legend.spacing.y = unit(.25, "cm")
  ) +
  guides(fill = guide_legend(byrow = TRUE)) +
  facet_wrap(vars(year), ncol = 4)
```


```{r greymulti, include=FALSE}
jawbone_all <- jawbone_all |>
  group_by(summer) |>
  mutate(order = row_number()) |>
  ungroup()

jawbone_2014 <- jawbone_all |>
  filter(year > syear) |>
  group_by(summer) |>
  mutate(order = row_number()) |>
  ungroup()

# Multi-year plot with current year overlaid as a heavier blue line
df  <- jawbone_2014
p2a <- ggplot(df, aes(x = day_plot, y = sst, group = year_chr)) +
  geom_line(colour = "grey60", linewidth = 0.25) +
  geom_line(data = df |> filter(year > pyear), aes(x = day_plot, y = sst), color = "blue", linewidth = 1) +
  scale_y_continuous(breaks = seq(tmin, tmax, 1)) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(y = "C", x = NULL, title = "Daily surface temperature, since 2014") +
  theme_minimal() +
  theme(legend.title = element_blank())

df  <- jawbone_all
p2b <- ggplot(df, aes(x = day_plot, y = sst, group = year_chr)) +
  geom_line(colour = "grey60", linewidth = 0.25) +
  geom_line(data = df |> filter(year > pyear), aes(x = day_plot, y = sst), color = "blue", linewidth = 1) +
  scale_y_continuous(breaks = seq(tmin, tmax, 1)) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(y = "C", x = NULL, title = "Daily surface temperature, since 2002") +
  theme_minimal() +
  theme(legend.title = element_blank())

df  <- jawbone_all |> filter(year > syear)
p2e <- ggplot(df, aes(x = order, y = sst, group = summer)) +
  geom_line(colour = "grey60", linewidth = 0.25) +
  geom_line(data = df |> filter(summer > pyear), aes(x = order, y = sst), color = "blue", linewidth = 1) +
  scale_y_continuous(breaks = seq(tmin, tmax, 1)) +
  scale_x_continuous(breaks = x_break, labels = x_label) +
  labs(y = "C", x = NULL, title = "Daily surface temperature, since 2014") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90), legend.title = element_blank())

df  <- jawbone_all |> filter(year > min(jawbone_all$year))
p2f <- ggplot(data = df, aes(x = order, y = sst, group = summer)) +
  geom_line(colour = "grey60", linewidth = 0.25) +
  geom_line(data = df |> filter(summer > pyear), aes(x = order, y = sst), color = "blue", linewidth = 1) +
  scale_y_continuous(breaks = seq(tmin, tmax, 1)) +
  scale_x_continuous(breaks = x_break, labels = x_label) +
  labs(y = "C", x = NULL, title = "Daily surface temperature, since 2003") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90), legend.title = element_blank())
```

```{r}
# Choose plot to do based on time of year; use Jan-Dec from April to Aug, Jul-June at other times of year
if(month(latest_time)>3 & month(latest_time)<9) {
  p2a
} else {
  p2e
}
if(month(latest_time)>3 & month(latest_time)<9) {
  p2b
} else {
  p2f
}

#p2c
```


```{r vsmean}
#Plot instead using averages for whole period or from syear onwards
l10y <- bay |>
  filter(year >= syear, longitude == wlon, latitude == wlat, year <= pyear) |>
  group_by(day_plot) |>
  mutate(Mean10 = mean(sst)) |>
  ungroup()

l20y <- bay |>
  filter(longitude == wlon, latitude == wlat, year <= pyear) |>
  group_by(day_plot) |>
  mutate(Mean20 = mean(sst)) |>
  ungroup()

p2d <- ggplot(data = jawbone_all, aes(x = day_plot, y = sst, group = year_chr)) +
#  geom_line(colour = "grey60", linewidth = 0.25) +
  geom_line(data = jawbone_all |> filter(year > pyear), aes(x = day_plot, y = sst), color = "blue", linewidth = 1) +
  geom_line(data = l10y, aes(x = day_plot, y = Mean10), color = "red", linewidth = 0.5) +
  geom_line(data = l20y, aes(x = day_plot, y = Mean20), color = "orange", linewidth = 0.5) +
  scale_y_continuous(breaks = seq(tmin, tmax, 1)) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(
    y        = "C",
    x        = NULL,
    title    = "Daily surface temperature",
    subtitle = "Averages since 2014 (red) and since 2002 (orange)"
  ) +
  theme_minimal() +
  theme(legend.title = element_blank())
p2d

```

## Three places of interest to divers and snorkellers

Open coast sites such as Flinders do not reach the same temperature extremes as those near the head of Port Phillip Bay (Williamstown). Sites near the bay entrance (Blairgowrie) are intermediate, being closer to open ocean conditions.

```{r threesite}
# Filter each location for the desired year, add a site label, then combine
bay3 <- bind_rows(
  bay |> filter(longitude == wlon, latitude == wlat, year == pyear) |> mutate(site = "Willy"),
  bay |> filter(longitude == blon, latitude == blat, year == pyear) |> mutate(site = "Blair"),
  bay |> filter(longitude == flon, latitude == flat, year == pyear) |> mutate(site = "Flinders")
)
```


```{r 3siteplotpen, include=FALSE}
# Now do multi-site plot
p2023 <- ggplot(data = bay3, aes(x = day_plot, y = sst, group = site, colour = site)) + 
  geom_line() + 
  scale_y_continuous(breaks = seq(tmin, tmax, 1), limits = c(tmin, tmax)) + 
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  scale_colour_uchicago() +
  theme_minimal() + 
  labs (title = "Daily surface temperature, last year",
        subtitle=pyear, y = "C", x = NULL) + 
  theme(legend.position = c(.7,.9), legend.title = element_blank()
  )
p2023
```


```{r 3siteplotcurr, include=FALSE}
# Do three place time series current year
# To do a different year, redefine the variable nyear earlier in code to set desired year
bay3 <- bind_rows(
  bay |> filter(longitude == wlon, latitude == wlat, year == nyear) |> mutate(site = "Willy"),
  bay |> filter(longitude == blon, latitude == blat, year == nyear) |> mutate(site = "Blair"),
  bay |> filter(longitude == flon, latitude == flat, year == nyear) |> mutate(site = "Flinders")
)
pnewyear <- ggplot(data = bay3, aes(x = day_plot, y = sst, group = site, colour = site)) + 
  geom_line() + 
#  geom_point(data = subset(bay3, time == max(time)), aes(x = day_plot, y = sst, group = site, colour = site)) +
  scale_y_continuous(breaks = seq(tmin, tmax, 1), limits = c(tmin, tmax)) + 
  scale_x_date(date_breaks = "2 month", date_labels = "%b", limits = as_date(c("2020-01-01","2020-12-31"))) +
  scale_colour_uchicago() +
  theme_minimal() + 
  labs (subtitle=nyear, y = "C", x = NULL) + 
  theme(legend.position = c(.7,.9), legend.title = element_blank())
pnewyear

```

```{r 3sitecurrvspastdec}
bay3 <- bind_rows(
  bay |> filter(longitude == wlon, latitude == wlat, year >= syear, year <= pyear) |> mutate(site = "Willy"),
  bay |> filter(longitude == blon, latitude == blat, year >= syear, year <= pyear) |> mutate(site = "Blair"),
  bay |> filter(longitude == flon, latitude == flat, year >= syear, year <= pyear) |> mutate(site = "Flinders")
) |>
  group_by(day_plot, site) |>
  mutate(Mean10 = mean(sst)) |>
  ungroup()

pdec <- ggplot(data = bay3, aes(x=day_plot, y=Mean10, group = site, colour = site)) + 
  geom_line() + 
  scale_y_continuous(breaks = seq(tmin, tmax, 1), limits = c(tmin, tmax)) + 
  scale_x_date(date_breaks = "2 month", date_labels = "%b") +
  scale_colour_uchicago() +
  theme_minimal() + 
  labs (subtitle="Past decade", y = "C", x = NULL) + 
  theme(legend.position = c(.7,.9), legend.title = element_blank())

```

```{r}
pnewyear + pdec + plot_layout(guides = 'collect')
```

```{r savefigs, include=FALSE, eval=FALSE}
# Save figures to file
ph <- 7   #Set height of exported pdf figure
pw <- 14  #Set width of exported pdf figure
ggsave("Jawbone.pdf", plot = p1, height = ph, width = pw, units='cm')
ggsave("Jawbone_multi.pdf", plot = p2, height = ph, width = pw, units='cm')
ggsave("Jawbone_multi2.pdf", plot = p2a, height = ph, width = pw, units='cm')
ggsave("3sites2023.pdf", plot = p2023, height = ph, width = pw, units='cm')
ggsave("3sites2024.pdf", plot = pnewyear, height = ph, width = pw, units='cm')

```

### Temperature contour map

```{r  tempcontour, warning=FALSE}
#Next two lines load coastline data for map plot. Easier to save to folder, and periodically run to check for updates
#coast <- gisco_get_coastallines(resolution = 1, cache = TRUE, update_cache = TRUE)
#save(coast,file = "coast.Rdata")
load("coast.Rdata")
ggplot() +
  geom_contour_filled(data = bay |> filter(date == latest_time), aes(x = longitude, y = latitude, z = sst)) +
  scale_fill_viridis_d(name = "Temp (C)", option = "B", direction = -1, begin = 0.4, end = 0.9) +
  geom_sf(data = coast, fill = "light grey") +
  coord_sf(
    xlim = lonx,
    ylim = latx
  )
```
