---
title: "Port Phillip Bay seawater temps"
output:
  html_document: 
       theme: cosmo
  pdf_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r include=FALSE}
library(tidyverse)
library(ggsci)
library(rerddap)
library(giscoR)
library(patchwork)
```


The temperature records are satellite sea surface temperatures. They are provided in several places, and what's here comes from
[NOAA](https://upwell.pfeg.noaa.gov/erddap/griddap/nesdisGeoPolarSSTN5NRT.html), presented as °C in the dataset *NOAA_DHW*. They come from grids 0.05° x 0.05°, corresponding to 5km x 5km. They can be viewed interactively, but it's not easy to look at multi-year time series or site comparisons.

The closest grid cell to Williamstown beach is centred on Lon. 144.875, Lat. -37.875. Two other snorkelling/diving spots are included; the SE corner of the bay (Blairgowrie, Rye, etc.) is 144.775, -38.375, and just outside Western Port (near Flinders pier) is 145.025, -38.475.

```{r params, include=FALSE}
#Set basic parameters for data download
sstInfo <- info('NOAA_DHW', url = "https://upwell.pfeg.noaa.gov/erddap/")
#Retrieves basic structure of NOAA dataset
#Downloading using erddap protocol
latx = c(-37.8, -38.525)    #Set bounding box for download
lonx = c(144.35, 145.3)

```

```{r include=FALSE, eval=FALSE}
#Use this chunk to make new clean copy of stored dataframe "bay"
#Normally just update saved file each time
timex = c("2003-01-01", "last")     #Start download from 2013
bay <- griddap(sstInfo, latitude = latx, longitude = lonx, time = timex, fmt= "csv", fields = 'CRW_SST')
save(bay, file="bay.Rdata")  #Saved version from last run
```

```{r getstart, include=FALSE}
load("bay.Rdata")     #Load saved file
start <- as.Date(max(bay$time))+1   #Get next day after data in saved file
```

```{r getdata, include=FALSE}
timex = c(as.character(start), "last")   #Used to download only new data, rather than large data set each time
baynew <- griddap(sstInfo, latitude = latx, longitude = lonx, time = timex, fmt= "csv", fields = 'CRW_SST')
bay <- rbind(bay, baynew)    #Append to saved file
bay <- unique(bay)
save(bay, file="bay.Rdata")  #Save new version
```


```{r tidy, include=FALSE }
bay$sst <- bay$crw_sst
bay$date <- ymd_hms(bay$time)
with_tz(bay$date, tzone="Australia/Melbourne")    #Convert to UTC+10
bay$year <- year(bay$date)   #Create year variable to allow grouping by year on plot
bay$year_chr <- as.character(bay$year)    #Produces better multi-year plot
bay$day <- yday(bay$date)   #Get day of year for each date
bay$day_plot <- as.Date("2019-12-31") + yday(bay$date)    #Creates variable amenable to m/d plots
bay$summer <- bay$year + (month(bay$date)>=7)      #Creates grouping by summer
bay$myLabel = case_when(day(bay$day_plot) == 1L ~ format(bay$day_plot, "%b-%d"),
                    TRUE ~ NA_character_)   #Adds label for 1st of month
latest_time <- max(bay$date)
latest<-date(latest_time)
```


```{r boundaries, include=FALSE}
#For convenience, set 3 desired locations as variables:
wlon <- 144.875
wlat <- -37.875
blon <- 144.775
blat <- -38.375
flon <- 145.025
flat <- -38.475

#Define year to be plotted and temperature ranges for y axis
# For Port Phillip Bay, force plots to useful temperature range (uses all of y-axis). 
# For viewing data from a single specific year, set value of pyear, rather than using formula below.
nyear <- max(bay$year)    #Current year in downloaded data
pyear <- nyear-1    #Set penultimate year
syear <- 2014  #Starting year for plots
tmax <- ceiling(max(bay$sst, na.rm = TRUE))
tmin <- floor(min(bay$sst, na.rm = TRUE))
wlatest <- subset(bay, time == max(bay$time) & longitude == wlon & latitude == wlat )
wtemp <- paste(format(latest,'%A, %B %d, %Y'), round(wlatest$sst, digits = 1), sep = ": ")

```

## Williamstown time series

#### Current year, with last year for an immediate comparison

```{r jawbonerecent, include=FALSE}
p1 <- ggplot(data = subset(bay, year >= pyear & longitude == wlon & latitude == wlat), aes(x=day_plot, y=sst, group=year_chr, colour=year_chr))+
  geom_line() +
  geom_point(data = subset(bay, year > pyear & longitude == wlon & latitude == wlat & time == max(time)), aes(x=day_plot, y=sst)) +
  scale_y_continuous(breaks=seq(tmin,tmax,1), limits = c(tmin,tmax)) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(y = "C", x = NULL, title = "Sea surface temperature, Williamstown", caption = wtemp) +
  theme_minimal() + 
  scale_color_aaas(labels=c("Last year", "This year")) +
  theme(legend.title=element_blank())
```

```{r jawbonesummer, include=FALSE}
df <- subset(bay, summer >= pyear & longitude == wlon & latitude == wlat)
df <- df %>% 
  group_by(summer) %>% 
  mutate(order = row_number()) %>% 
  ungroup()
x_break <- df$order[!is.na(df$myLabel)][1:12]
x_label <- df$myLabel[x_break]
#x_label
p1a <- ggplot(data = df, aes(x=order, y=sst, group=as.factor(summer), colour=as.factor(summer)))+
  geom_line() +
  geom_point(data = subset(df, summer > pyear & time == max(time)), aes(x=order, y=sst)) +
  scale_y_continuous(breaks=seq(tmin,tmax,1), limits = c(tmin,tmax)) +
  scale_x_continuous(
    breaks = x_break,
    labels = x_label) +
  labs(y = "C", x = NULL, title = "Sea surface temperature, Williamstown", caption = wtemp) +
  theme_minimal() + 
  scale_color_aaas(labels=c("Last summer","This summer")) +
  theme(axis.text.x = element_text(angle = 90),legend.title=element_blank())
```

```{r}
# Choose plot to do based on time of year; use Jan-Dec from April to Aug, Jul-June at other times of year
if(month(latest_time)>3 & month(latest_time)<9) {
  p1
} else {
  p1a
}
```

#### Multi-year plots

The first figure shows the current year, other years from 2014 onwards in the background, and the second includes the past 22 years. The third figure shows the current year against 10- and 20-year averages.

```{r multipanel, include=FALSE}
#Data goes from 2002 to present. Having 23 years' data is a bit spaghetti-like. Two options are to plot fewer years and/or emphasise the current year more. Do the first option by specifying a value for syear.
#Do multi-year plot, then overlay data for chosen year (pyear) with heavier line.
jawbone_all <- subset(bay, longitude == wlon & latitude == wlat)
p2<- ggplot(data = subset(jawbone_all, year >= syear), aes(x=day_plot, y=sst))+
  geom_line(linewidth=0.25) +
  scale_y_continuous(breaks=seq(tmin,tmax,2)) +
  scale_x_date(date_breaks = "2 month", date_labels = "%b") +
  labs(y = "C", x = NULL, title = "Daily surface temperature each year") +
  theme_classic()+
  theme(legend.title=element_blank(),
        legend.text = element_text(size = 5),
        theme(legend.spacing.y = unit(.25, 'cm'))
        ) +
  guides(fill = guide_legend(byrow = TRUE)) +
  facet_wrap(vars(year))

p2c<- ggplot(jawbone_all, aes(x=day_plot, y=sst))+
  geom_line(linewidth=0.25) +
  scale_y_continuous(breaks=seq(tmin,tmax,2)) +
  scale_x_date(date_breaks = "2 month", date_labels = "%b") +
  labs(y = "C", x = NULL, title = "Daily surface temperature each year") +
  theme_classic()+
  theme(legend.title=element_blank(),
        legend.text = element_text(size = 5),
        theme(legend.spacing.y = unit(.25, 'cm'))
        ) +
  guides(fill = guide_legend(byrow = TRUE)) +
  facet_wrap(vars(year), ncol=4)
```


```{r greymulti, include=FALSE}
jawbone_all <- jawbone_all %>% 
  group_by(summer) %>% 
  mutate(order = row_number()) %>% 
  ungroup()
#x_break <- df$order[!is.na(df$myLabel)][1:12]
#x_label <- df$myLabel[x_break]
#x_label
jawbone_2014 <- subset(jawbone_all, year > syear)
jawbone_2014 <- jawbone_2014 %>% 
  group_by(summer) %>% 
  mutate(order = row_number()) %>% 
  ungroup()

#Do multi-year plot, then overlay data for chosen year (pyear) with heavier black line.
df <- jawbone_2014
p2a<-ggplot(df, aes(x=day_plot, y=sst, group=year_chr))+
  geom_line( colour="grey60", linewidth=0.25) +
  geom_line(data = subset(df, year > pyear), aes(x=day_plot, y=sst), color="blue", linewidth=1) +
  scale_y_continuous(breaks=seq(tmin,tmax,1)) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(y = "C", x = NULL, title = "Daily surface temperature, since 2014") +
  theme_minimal() + 
  theme(legend.title=element_blank())

df <- jawbone_all
p2b<-ggplot(df, aes(x=day_plot, y=sst, group=year_chr))+
  geom_line( colour="grey60", linewidth=0.25) +
  geom_line(data = subset(df, year > pyear), aes(x=day_plot, y=sst), color="blue", linewidth=1) +
  scale_y_continuous(breaks=seq(tmin,tmax,1)) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(y = "C", x = NULL, title = "Daily surface temperature, since 2002") +
  theme_minimal() + 
  theme(legend.title=element_blank())

df <- subset(jawbone_all, year > syear)
p2e <- ggplot(df, aes(x=order, y=sst, group=summer))+
  geom_line( colour="grey60", linewidth=0.25) +
  geom_line(data = subset(df, summer > pyear), aes(x=order, y=sst), color="blue", linewidth=1) +
  scale_y_continuous(breaks=seq(tmin,tmax,1)) +
  scale_x_continuous(
    breaks = x_break,
    labels = x_label) +
  labs(y = "C", x = NULL, title = "Daily surface temperature, since 2014") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90),legend.title=element_blank())

df<- subset(jawbone_all, year > min(jawbone_all$year))
p2f <- ggplot(data = df, aes(x=order, y=sst, group=summer))+
  geom_line( colour="grey60", linewidth=0.25) +
  geom_line(data = subset(df, summer > pyear), aes(x=order, y=sst), color="blue", linewidth=1) +
  scale_y_continuous(breaks=seq(tmin,tmax,1)) +
  scale_x_continuous(
    breaks = x_break,
    labels = x_label) +
  labs(y = "C", x = NULL, title = "Daily surface temperature, since 2003") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90),legend.title=element_blank())
```

```{r}
# Choose plot to do based on time of year; use Jan-Dec from April to Aug, Jul-June at other times of year
if(month(latest_time)>3 & month(latest_time)<9) {
  p2a
} else {
  p2e
}
if(month(latest_time)>3 & month(latest_time)<9) {
  p2b
} else {
  p2f
}

#p2c
```


```{r vsmean}
#Plot instead using averages for whole period or from syear onwards
l10y<-subset(bay, year >= syear & longitude == wlon & latitude == wlat & year<=pyear)
l10y <- l10y %>%
    group_by(day_plot) %>%
    mutate(Mean10 = mean(sst)) %>%
    ungroup
l20y<-subset(bay, longitude == wlon & latitude == wlat & year<=pyear)
l20y <- l20y %>%
    group_by(day_plot) %>%
    mutate(Mean20 = mean(sst)) %>%
    ungroup
p2d<-ggplot(data = jawbone_all, aes(x=day_plot, y=sst, group=year_chr))+
#  geom_line( colour="grey60", linewidth=0.25) +
  geom_line(data = subset(jawbone_all, year > pyear), aes(x=day_plot, y=sst), color="blue", linewidth=1) +
  geom_line(data=l10y, aes(x=day_plot, y=Mean10), color="red", linewidth=0.5)+
  geom_line(data=l20y, aes(x=day_plot, y=Mean20), color= "orange", linewidth=0.5)+
  scale_y_continuous(breaks=seq(tmin,tmax,1)) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(y = "C", x = NULL, title = "Daily surface temperature",
       subtitle="Averages since 2014 (red) and since 2002 (orange)") +
  theme_minimal() + 
  theme(legend.title=element_blank())
#p2a 
#p2b
p2d

```

## Three places of interest to divers and snorkellers

Open coast sites, such as Flinders, don't get as warm or as cold as those near the top of Port Phillip Bay (Williamstown). Places near the entrance of the bay (Blairgowrie) are intermediate, but closer to open ocean.

```{r threesite}
# Use 3 initial data sets, jb, bl, fl, convert them to CSV dataframes,then join them end to end. Add a variable for sites
# First file is Jawbone
jb <- subset(bay, longitude == wlon & latitude == wlat & year == pyear) 
jb$site <- "Willy"

# Now blair 
bl <- subset(bay, longitude == blon & latitude == blat & year == pyear)
bl$site <- "Blair"

# Now Flinders
fl <- subset(bay, longitude == flon & latitude == flat & year == pyear) 
fl$site <- "Flinders" 

# Merge sites 
bay3 <-do.call("rbind", list(jb, bl, fl)) 
```


```{r 3siteplotpen, include=FALSE}
# Now do multi-site plot
p2023 <- ggplot(data = bay3, aes(x=day_plot, y=sst, group = site, colour = site)) + 
  geom_line() + 
  scale_y_continuous(breaks=seq(tmin,tmax,1), limits = c(tmin,tmax)) + 
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  scale_colour_uchicago() +
  theme_minimal() + 
  labs (title = "Daily surface temperature, last year",
        subtitle=pyear, y = "C", x = NULL) + 
  theme(legend.position = c(.7,.9), legend.title = element_blank())
p2023
```


```{r 3siteplotcurr, include=FALSE}
# Do three place time series current year
# To do a different year, redefine the variable nyear earlier in code to set desired year
jb <- subset(bay, longitude == wlon & latitude == wlat & year == nyear) 
jb$site <- "Willy"
bl <- subset(bay, longitude == blon & latitude == blat & year == nyear)
bl$site <- "Blair"
fl <- subset(bay, longitude == flon & latitude == flat & year == nyear) 
fl$site <- "Flinders" 
bay3 <-do.call("rbind", list(jb, bl, fl)) 
pnewyear <- ggplot(data = bay3, aes(x=day_plot, y=sst, group = site, colour = site)) + 
  geom_line() + 
#  geom_point(data = subset(bay3, time == max(time)), aes(x=day_plot, y=sst, group = site, colour = site)) +
  scale_y_continuous(breaks=seq(tmin,tmax,1), limits = c(tmin,tmax)) + 
  scale_x_date(date_breaks = "2 month", date_labels = "%b", limits = as_date(c("2020-01-01","2020-12-31"))) +
  scale_colour_uchicago() +
  theme_minimal() + 
  labs (subtitle=nyear, y = "C", x = NULL) + 
  theme(legend.position = c(.7,.9), legend.title = element_blank())
pnewyear

```

```{r 3sitecurrvspastdec}
jb <- subset(bay, longitude == wlon & latitude == wlat & year >= syear & year <=pyear) 
jb$site <- "Willy"
bl <- subset(bay, longitude == blon & latitude == blat & year >= syear & year <=pyear)
bl$site <- "Blair"
fl <- subset(bay, longitude == flon & latitude == flat & year >= syear & year <=pyear) 
fl$site <- "Flinders" 
bay3 <-do.call("rbind", list(jb, bl, fl)) 

#l10y<-subset(bay, year >= syear & year<=pyear)
bay3 <- bay3 %>%
    group_by(day_plot, site) %>%
    mutate(Mean10 = mean(sst)) %>%
    ungroup

pdec <- ggplot(data = bay3, aes(x=day_plot, y=Mean10, group = site, colour = site)) + 
  geom_line() + 
  scale_y_continuous(breaks=seq(tmin,tmax,1), limits = c(tmin,tmax)) + 
  scale_x_date(date_breaks = "2 month", date_labels = "%b") +
  scale_colour_uchicago() +
  theme_minimal() + 
  labs (subtitle="Past decade", y = "C", x = NULL) + 
  theme(legend.position = c(.7,.9), legend.title = element_blank())

```

```{r}
pnewyear+pdec +plot_layout(guides='collect')
```

```{r savefigs, include=FALSE, eval=FALSE}
# Save figures to file
ph <- 7   #Set height of exported pdf figure
pw <- 14  #Set width of exported pdf figure
ggsave("Jawbone.pdf", plot = p1, height = ph, width = pw, units='cm')
ggsave("Jawbone_multi.pdf", plot = p2, height = ph, width = pw, units='cm')
ggsave("Jawbone_multi2.pdf", plot = p2a, height = ph, width = pw, units='cm')
ggsave("3sites2023.pdf", plot = p2023, height = ph, width = pw, units='cm')
ggsave("3sites2024.pdf", plot = pnewyear, height = ph, width = pw, units='cm')

```

### Temperature contour map

```{r  tempcontour, warning=FALSE}
#Next two lines load coastline data for map plot. Easier to save to folder, and periodically run to check for updates
#coast <- gisco_get_coastallines(resolution = 1, cache = TRUE, update_cache = TRUE)
#save(coast,file = "coast.Rdata")
load("coast.Rdata")
ggplot() +
  geom_contour_filled(data = subset(bay, date==latest_time), aes(x=longitude, y=latitude,z=sst)) +
  scale_fill_viridis_d(name = "Temp (C)", option = "B", direction = -1, begin = 0.4, end = 0.9) +
  geom_sf(data = coast, fill = "light grey") +
  coord_sf(
    xlim = lonx,
    ylim = latx
  )
```
