---
title: "Port Phillip Bay seawater temps"
output:
  html_document: 
       theme: cosmo
  pdf_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r loadlib, include=FALSE}
library(tidyverse)
library(ggsci)
library(rerddap)
library(giscoR)
```


The temperature records used here are satellite sea surface temperatures. They are provided in several places, and what's here comes from
[NOAA](https://pae-paha.pacioos.hawaii.edu/erddap/griddap/dhw_5km.html), presented as °C in the dataset *dhw_5km*. This dataset has data from grids 0.05° x 0.05°, corresponding to 5km x 5km.

The closest grid cell to Williamstown beach is centred on Lon. 144.875, Lat. -37.875. Two other snorkelling/diving spots are included; the SE corner of the bay (Blairgowrie, Rye, etc.) is 144.775, -38.375, and just outside Western Port (near Flinders pier) is 145.025, -38.475.

The temperatures are daily measurements, so the graphs are a bit noisy; some other places show temperatures averaged over a few days to smooth things out a bit.

```{r params, include=FALSE}
#Set basic parameters for data download
#Downloading using erddap protocol
sstInfo <- info('dhw_5km', url = "https://pae-paha.pacioos.hawaii.edu/erddap/")
#Retrieves basic structure of NOAA dataset
```

```{r boundbox, include=FALSE}
latx = c(-37.8, -38.525)    #Set bounding box for download
lonx = c(144.35, 145.3)
```

```{r cleancopy, include=FALSE, eval=FALSE}
#Use this chunk to make new clean copy of stored dataframe "bay"
#Normally just update saved file each time
timex = c("2003-01-01", "last")     #Start download from 2013
bay <- griddap(sstInfo, latitude = latx, longitude = lonx, time = timex, fmt= "csv", fields = 'crw_sst')
save(bay, file="bay.Rdata")  #Saved version from last run
```

```{r getstart, include=FALSE}
load("bay.Rdata")     #Load saved file
start <- as.Date(max(bay$time))+1   #Get next day after data in saved file
```

```{r getdata, include=FALSE}
timex = c(as.character(start), "last")   #Used to download only new data, rather than large data set each time
baynew <- griddap(sstInfo, latitude = latx, longitude = lonx, time = timex, fmt= "csv", fields = 'CRW_SST')   #Check; new mirror may use lower case crw...
bay <- bind_rows(bay, baynew)  # Append to saved file
bay <- unique(bay)
save(bay, file="bay.Rdata")  #Save new version
```


```{r tidy, include=FALSE}
bay <- bay |>
  mutate(
    sst      = crw_sst,
    date     = with_tz(ymd_hms(time), tzone = "Australia/Melbourne"),  # Convert to UTC+10
    year     = year(date),                          # Year variable for grouping
    year_chr = as.character(year),                  # Character year for multi-year plots
    day      = yday(date),                          # Day of year
    day_plot = as.Date("2019-12-31") + yday(date),  # Variable for m/d plots
    summer   = year + (month(date) >= 7),           # Grouping by summer
    myLabel  = case_when(                           # Label for 1st of each month
      day(day_plot) == 1L ~ format(day_plot, "%b-%d"),
      TRUE               ~ NA_character_
    )
  )
latest_time <- max(bay$date)
latest      <- date(latest_time)
```


```{r boundaries, include=FALSE}
#For convenience, set 3 desired locations as variables:
wlon <- 144.875
wlat <- -37.875
blon <- 144.775
blat <- -38.375
flon <- 145.025
flat <- -38.475

#Define year to be plotted and temperature ranges for y axis
# For Port Phillip Bay, force plots to useful temperature range (uses all of y-axis). 
# For viewing data from a single specific year, set value of pyear, rather than using formula below.
nyear <- max(bay$year)    #Current year in downloaded data
pyear <- nyear-1    #Set penultimate year
syear <- 2014  #Starting year for plots
tmax <- ceiling(max(bay$sst, na.rm = TRUE))
tmin <- floor(min(bay$sst, na.rm = TRUE))
wlatest <- bay |> filter(time == max(time), longitude == wlon, latitude == wlat)
wtemp   <- paste(format(latest, "%A, %B %d, %Y"), round(wlatest$sst, digits = 1), sep = ": ")

```

## Williamstown time series

#### Current year, with last year for an immediate comparison

```{r jawbonerecent, include=FALSE}
p1 <- ggplot(
    data = bay |> filter(year >= pyear, longitude == wlon, latitude == wlat),
    aes(x = day_plot, y = sst, group = year_chr, colour = year_chr)
  ) +
  geom_line() +
  geom_point(
    data = bay |> filter(year > pyear, longitude == wlon, latitude == wlat, time == max(time)),
    aes(x = day_plot, y = sst)
  ) +
  scale_y_continuous(breaks = seq(tmin, tmax, 1), limits = c(tmin, tmax)) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(y = "C", x = NULL, title = "Sea surface temperature, Williamstown", caption = wtemp) +
  theme_minimal() +
  scale_color_aaas(labels = c("Last year", "This year")) +
  theme(legend.title = element_blank())
```

```{r jawbonesummer, include=FALSE}
df <- bay |>
  filter(summer >= pyear, longitude == wlon, latitude == wlat) |>
  group_by(summer) |>
  mutate(order = row_number()) |>
  ungroup()
x_break <- df$order[!is.na(df$myLabel)][1:12]
x_label <- df$myLabel[x_break]
p1a <- ggplot(data = df, aes(x = order, y = sst, group = as.factor(summer), colour = as.factor(summer))) +
  geom_line() +
  geom_point(
    data = df |> filter(summer > pyear, time == max(time)),
    aes(x = order, y = sst)
  ) +
  scale_y_continuous(breaks = seq(tmin, tmax, 1), limits = c(tmin, tmax)) +
  scale_x_continuous(breaks = x_break, labels = x_label) +
  labs(y = "C", x = NULL, title = "Sea surface temperature, Williamstown", caption = wtemp) +
  theme_minimal() +
  scale_color_aaas(labels = c("Last summer", "This summer")) +
  theme(axis.text.x = element_text(angle = 90), legend.title = element_blank())
```

```{r}
# Choose plot to do based on time of year; use Jan-Dec from April to Aug, Jul-June at other times of year
if(month(latest_time)>3 & month(latest_time)<9) {
  p1
} else {
  p1a
}
```

#### Multi-year plots

The first figure shows the current year, other years from 2014 onwards in the background, and the second includes the past 22 years. The third figure shows the current year against 10- and 20-year averages.

```{r multipanel, include=FALSE}
#Data goes from 2002 to present. Having 23 years' data is a bit spaghetti-like. Two options are to plot fewer years and/or emphasise the current year more. Do the first option by specifying a value for syear.
#Do multi-year plot, then overlay data for chosen year (pyear) with heavier line.
jawbone_all <- bay |> filter(longitude == wlon, latitude == wlat)
p2 <- ggplot(data = jawbone_all |> filter(year >= syear), aes(x = day_plot, y = sst)) +
  geom_line(linewidth = 0.25) +
  scale_y_continuous(breaks = seq(tmin, tmax, 2)) +
  scale_x_date(date_breaks = "2 month", date_labels = "%b") +
  labs(y = "C", x = NULL, title = "Daily surface temperature each year") +
  theme_classic() +
  theme(
    legend.title     = element_blank(),
    legend.text      = element_text(size = 5),
    legend.spacing.y = unit(.25, "cm")
  ) +
  guides(fill = guide_legend(byrow = TRUE)) +
  facet_wrap(vars(year))

p2c <- ggplot(jawbone_all, aes(x = day_plot, y = sst)) +
  geom_line(linewidth = 0.25) +
  scale_y_continuous(breaks = seq(tmin, tmax, 2)) +
  scale_x_date(date_breaks = "2 month", date_labels = "%b") +
  labs(y = "C", x = NULL, title = "Daily surface temperature each year") +
  theme_classic() +
  theme(
    legend.title     = element_blank(),
    legend.text      = element_text(size = 5),
    legend.spacing.y = unit(.25, "cm")
  ) +
  guides(fill = guide_legend(byrow = TRUE)) +
  facet_wrap(vars(year), ncol = 4)
```


```{r greymulti, include=FALSE}
jawbone_all <- jawbone_all |>
  group_by(summer) |>
  mutate(order = row_number()) |>
  ungroup()

jawbone_2014 <- jawbone_all |>
  filter(year > syear) |>
  group_by(summer) |>
  mutate(order = row_number()) |>
  ungroup()

# Multi-year plot with current year overlaid as a heavier blue line
df  <- jawbone_2014
p2a <- ggplot(df, aes(x = day_plot, y = sst, group = year_chr)) +
  geom_line(colour = "grey60", linewidth = 0.25) +
  geom_line(data = df |> filter(year > pyear), aes(x = day_plot, y = sst), color = "blue", linewidth = 1) +
  scale_y_continuous(breaks = seq(tmin, tmax, 1)) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(y = "C", x = NULL, title = "Daily surface temperature, since 2014") +
  theme_minimal() +
  theme(legend.title = element_blank())

df  <- jawbone_all
p2b <- ggplot(df, aes(x = day_plot, y = sst, group = year_chr)) +
  geom_line(colour = "grey60", linewidth = 0.25) +
  geom_line(data = df |> filter(year > pyear), aes(x = day_plot, y = sst), color = "blue", linewidth = 1) +
  scale_y_continuous(breaks = seq(tmin, tmax, 1)) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(y = "C", x = NULL, title = "Daily surface temperature, since 2002") +
  theme_minimal() +
  theme(legend.title = element_blank())

df  <- jawbone_all |> filter(year > syear)
p2e <- ggplot(df, aes(x = order, y = sst, group = summer)) +
  geom_line(colour = "grey60", linewidth = 0.25) +
  geom_line(data = df |> filter(summer > pyear), aes(x = order, y = sst), color = "blue", linewidth = 1) +
  scale_y_continuous(breaks = seq(tmin, tmax, 1)) +
  scale_x_continuous(breaks = x_break, labels = x_label) +
  labs(y = "C", x = NULL, title = "Daily surface temperature, since 2014") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90), legend.title = element_blank())

df  <- jawbone_all |> filter(year > min(jawbone_all$year))
p2f <- ggplot(data = df, aes(x = order, y = sst, group = summer)) +
  geom_line(colour = "grey60", linewidth = 0.25) +
  geom_line(data = df |> filter(summer > pyear), aes(x = order, y = sst), color = "blue", linewidth = 1) +
  scale_y_continuous(breaks = seq(tmin, tmax, 1)) +
  scale_x_continuous(breaks = x_break, labels = x_label) +
  labs(y = "C", x = NULL, title = "Daily surface temperature, since 2003") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90), legend.title = element_blank())
```

```{r}
# Choose plot to do based on time of year; use Jan-Dec from April to Aug, Jul-June at other times of year
if(month(latest_time)>3 & month(latest_time)<9) {
  p2a
} else {
  p2e
}
if(month(latest_time)>3 & month(latest_time)<9) {
  p2b
} else {
  p2f
}

#p2c
```


```{r vsmean}
#Plot instead using averages for whole period or from syear onwards
l10y <- bay |>
  filter(year >= syear, longitude == wlon, latitude == wlat, year <= pyear) |>
  group_by(day_plot) |>
  mutate(Mean10 = mean(sst)) |>
  ungroup()

l20y <- bay |>
  filter(longitude == wlon, latitude == wlat, year <= pyear) |>
  group_by(day_plot) |>
  mutate(Mean20 = mean(sst)) |>
  ungroup()

p2d <- ggplot(data = jawbone_all, aes(x = day_plot, y = sst, group = year_chr)) +
#  geom_line(colour = "grey60", linewidth = 0.25) +
  geom_line(data = jawbone_all |> filter(year > pyear), aes(x = day_plot, y = sst), color = "blue", linewidth = 1) +
  geom_line(data = l10y, aes(x = day_plot, y = Mean10), color = "red", linewidth = 0.5) +
  geom_line(data = l20y, aes(x = day_plot, y = Mean20), color = "orange", linewidth = 0.5) +
  scale_y_continuous(breaks = seq(tmin, tmax, 1)) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(
    y        = "C",
    x        = NULL,
    title    = "Daily surface temperature",
    subtitle = "Averages since 2014 (red) and since 2002 (orange)"
  ) +
  theme_minimal() +
  theme(legend.title = element_blank())
p2d

```

## Three places of interest to divers and snorkellers

Open coast sites, such as Flinders, don't get as warm or as cold as those near the top of Port Phillip Bay (Williamstown). Places near the entrance of the bay (Blairgowrie) are intermediate, but closer to open ocean conditions.

```{r threesite}
# Filter each location for the desired year, add a site label, then combine
bay3 <- bind_rows(
  bay |> filter(longitude == wlon, latitude == wlat, year == pyear) |> mutate(site = "Willy"),
  bay |> filter(longitude == blon, latitude == blat, year == pyear) |> mutate(site = "Blair"),
  bay |> filter(longitude == flon, latitude == flat, year == pyear) |> mutate(site = "Flinders")
)
```


```{r 3siteplotpen, include=FALSE}
# Now do multi-site plot
p2023 <- ggplot(data = bay3, aes(x=day_plot, y=sst, group = site, colour = site)) + 
  geom_line() + 
  scale_y_continuous(breaks=seq(tmin,tmax,1), limits = c(tmin,tmax)) + 
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  scale_colour_uchicago() +
  theme_minimal() + 
  labs (title = "Daily surface temperature, last year",
        subtitle=pyear, y = "C", x = NULL) + 
  theme(legend.position = c(.7,.9), legend.title = element_blank())
p2023
```


```{r 3siteplotcurr, include=FALSE}
# Do three place time series current year
# To do a different year, redefine the variable nyear earlier in code to set desired year
bay3 <- bind_rows(
  bay |> filter(longitude == wlon, latitude == wlat, year == nyear) |> mutate(site = "Willy"),
  bay |> filter(longitude == blon, latitude == blat, year == nyear) |> mutate(site = "Blair"),
  bay |> filter(longitude == flon, latitude == flat, year == nyear) |> mutate(site = "Flinders")
)
pnewyear <- ggplot(data = bay3, aes(x=day_plot, y=sst, group = site, colour = site)) + 
  geom_line() + 
#  geom_point(data = subset(bay3, time == max(time)), aes(x=day_plot, y=sst, group = site, colour = site)) +
  scale_y_continuous(breaks=seq(tmin,tmax,1), limits = c(tmin,tmax)) + 
  scale_x_date(date_breaks = "2 month", date_labels = "%b", limits = as_date(c("2020-01-01","2020-12-31"))) +
  scale_colour_uchicago() +
  theme_minimal() + 
  labs (subtitle=nyear, y = "C", x = NULL) + 
  theme(legend.position = c(.7,.9), legend.title = element_blank())
pnewyear

```

```{r 3sitecurrvspastdec}
bay3 <- bind_rows(
  bay |> filter(longitude == wlon, latitude == wlat, year >= syear, year <= pyear) |> mutate(site = "Willy"),
  bay |> filter(longitude == blon, latitude == blat, year >= syear, year <= pyear) |> mutate(site = "Blair"),
  bay |> filter(longitude == flon, latitude == flat, year >= syear, year <= pyear) |> mutate(site = "Flinders")
) |>
  group_by(day_plot, site) |>
  mutate(Mean10 = mean(sst)) |>
  ungroup()

pdec <- ggplot(data = bay3, aes(x=day_plot, y=Mean10, group = site, colour = site)) + 
  geom_line() + 
  scale_y_continuous(breaks=seq(tmin,tmax,1), limits = c(tmin,tmax)) + 
  scale_x_date(date_breaks = "2 month", date_labels = "%b") +
  scale_colour_uchicago() +
  theme_minimal() + 
  labs (subtitle="Past decade", y = "C", x = NULL) + 
  theme(legend.position = c(.7,.9), legend.title = element_blank())

```

```{r}
pnewyear
pdec
```

```{r savefigs, include=FALSE, eval=FALSE}
# Save figures to file
ph <- 7   #Set height of exported pdf figure
pw <- 14  #Set width of exported pdf figure
ggsave("Jawbone.pdf", plot = p1, height = ph, width = pw, units='cm')
ggsave("Jawbone_multi.pdf", plot = p2, height = ph, width = pw, units='cm')
ggsave("Jawbone_multi2.pdf", plot = p2a, height = ph, width = pw, units='cm')
ggsave("3sites2023.pdf", plot = p2023, height = ph, width = pw, units='cm')
ggsave("3sites2024.pdf", plot = pnewyear, height = ph, width = pw, units='cm')

```

### Temperature contour map

```{r  tempcontour, warning=FALSE}
#Next two lines load coastline data for map plot. Easier to save to folder, and periodically run to check for updates
#coast <- gisco_get_coastallines(resolution = 1, cache = TRUE, update_cache = TRUE)
#save(coast,file = "coast.Rdata")
load("coast.Rdata")
ggplot() +
  geom_contour_filled(data = bay |> filter(date == latest_time), aes(x = longitude, y = latitude, z = sst)) +
  scale_fill_viridis_d(name = "Temp (C)", option = "B", direction = -1, begin = 0.4, end = 0.9) +
  geom_sf(data = coast, fill = "light grey") +
  coord_sf(
    xlim = lonx,
    ylim = latx
  )
```
